\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2025-05-30 misc/JSS article/final/jss5278/jss5278.tex
Submission package.

\documentclass[nojss]{jss} % article

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, including vectorized and verbose hash-joins and fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications of the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

@@

In summary, \pkg{collapse} is a new foundation package for advanced statistical computing and data transformation in \proglang{R} that integrates seamlessly with the ecosystem and offers outstanding speed and memory efficiency. Thus, many core tasks can be done with \pkg{collapse}, and easily extended by specialized packages, yielding more lightweight, faster, and shorter \proglang{R} programs. \newline
% TODO: Need examples of this

@@

While this article cannot fully present \pkg{collapse}, the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

>>>>

\documentclass[article]{jss} % article

@@

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844. \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides powerful data manipulation functions, including vectorized and verbose hash-joins and fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools. This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels. The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames, and seamlessly supporting popular extensions to these objects such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is also extensively globally and interactively configurable. \newline

@@

% In summary, \pkg{collapse} is a new foundation package for advanced statistical computing and data transformation in \proglang{R} that integrates seamlessly with the ecosystem and offers outstanding speed and memory efficiency. Thus, many core tasks can be done with \pkg{collapse}, and easily extended by specialized packages, yielding more lightweight, faster, and shorter \proglang{R} programs. \newline
% TODO: Need examples of this

@@

This article cannot fully present \pkg{collapse}, but the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
